Early Warning Signals and the Prosecutor's Fallacy

### 2012-09-04 prosecutor/fallacy.tex
copy sent to alan

@@

To demonstrate the effect we observe is not unique to models with Allee
effects, we provide an example of the effect arising in a discrete-time
model with two non-zero stable states adapted from~\citep{May1977},
\begin{equation}
X_{t+1} = X_t \exp\left( r \left(1 - \frac{ X_t }{ K } \right) - \frac{ a * X_t ^ {Q - 1} }{s ^ Q + H ^ Q} \right). \label{May1977}
\end{equation}
Annotated, parallel code to replicate the analysis presented here is
also provided in the appendix.

\begin{figure}
  \begin{center}
    \ifhavefigures
    \includegraphics[width=6in]{figure2}
    \fi
  \end{center}
  \caption{The distribution of the correlation statistic $\tau$ for two
  early warning indicators (variance, autocorrelation) on replicates
  conditionally selected for having collapsed by chance in simulations
  is shown in grey bars.  Solid lines indicate the estimated density of
  the statistic from a random sample of the simulations (not conditional
  on observing a transition). Positive values of $\tau$ correspond to
  a pattern of an indicator increasing with time; typically taken as
  evidence that a system is approaching a critical transition.  In these
  simulations, the pattern arises instead from the Prosecutor's fallacy
  of conditional selection.}
  \label{fig:indicator}
\end{figure}

@@

\section{References}%bibliography
\bibliography{fallacy}

>>>>

\usepackage[hyphens]{url}
\usepackage[pdftex]{color}
\usepackage[breaklinks=true,linktocpage,colorlinks]{hyperref}
\hypersetup{breaklinks=true, pdfborder={0 0 0}}

@@

To demonstrate the effect we observe is not unique to models with Allee
effects, we provide an example of the effect arising in a discrete-time
model with two non-zero stable states adapted from~\citep{May1977},
\begin{equation}
X_{t+1} = X_t \exp\left( r \left(1 - \frac{ X_t }{ K } \right) - \frac{ a * X_t ^ {Q - 1} }{X_t ^ Q + H ^ Q} \right). \label{May1977}
\end{equation}
which combines a logistic growth model with a saturating predator response (See~\citet{May1977} for detailed discussion), shown in Figure~\ref{fig:may}.  Code to replicate the analysis can be found at \\ 
\begin{sloppy}\href{https://github.com/cboettig/earlywarning/tree/prosecutor/}{https://github.com/cboettig/earlywarning/tree/prosecutor/}\end{sloppy}.

\begin{figure}
  \begin{center}
    \ifhavefigures
    \includegraphics[width=6in]{figure2}
    \fi
  \end{center}
  \caption{The distribution of the correlation statistic $\tau$ for two
  early warning indicators (variance, autocorrelation) on replicates
  conditionally selected for having collapsed by chance in simulations
  is shown in grey bars.  Solid lines indicate the estimated density of
  the statistic from a random sample of the simulations (not conditional
  on observing a transition). Positive values of $\tau$ correspond to
  a pattern of an indicator increasing with time; typically taken as
  evidence that a system is approaching a critical transition.  In these
  simulations, the pattern arises instead from the Prosecutor's fallacy
  of conditional selection. (the slight bimodality in the null distribution
  arises from limited sample size).}
  \label{fig:indicator}
\end{figure}

\begin{figure}
  \begin{center}
    \ifhavefigures
    \includegraphics[width=6in]{figure3}
    \fi
  \end{center}
  \caption{The identical analysis from Figure~\ref{fig:indicator} is shown for the model in Eq~\eqref{May1977} using parameters $r = 0.75$, $K=10$, $a=1.7$, $Q=3$, and $H=1$. A similar statistical bias, particularly towards positive values of $tau$ occurs in this model as well.}
  \label{fig:may}
\end{figure}

@@

\bibliography{fallacy}
